A Neurally-Guided, Parallel Theorem Prover*

### 2019-06-30 papers/FroCoS19/paper.tex
FroCoS '19 (partial) feedback

\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neural, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
	Giles Reger
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}, \email{giles.reger@manchester.ac.uk}
}
\maketitle

@@

Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.

@@

Processing structured data such as logical formulae is a relatively new domain for neural networks.
Some work attempts to use unstructured representations of such formulae, such as text, or build entirely-new models for a specific logic~\cite{logical-entailment}, whereas others attempt to re-use neural techniques for generic structures such as trees~\cite{logical-tree}.
A promising direction in this area is recent research on neural methods working with graphs~\cite{graph-cnn,gcn,gcn-relational}, which have already been applied to premise selection~\cite{formula-graph}.
Graph neural networks tend to include network layers inspired by convolution operators in image-processing networks, combining information from neighbouring nodes (pixels).
For example, the Graph Convolutional Network~\cite{gcn} includes the following convolution layer, where \(H_i\) is the network output at layer \(i\), \(\bar{A}\) is the self-connected adjacency matrix, \(\bar{D}\) is the diagonal node degree matrix of \(\bar{A}\), \(W_i\) is a trainable weight matrix and \(\sigma\) is a non-linear activation function:
\[
	H_{i + 1} = \sigma\left(\bar{D}^{-\frac{1}{2}}\bar{A}\bar{D}^{-\frac{1}{2}}H_iW_i\right)
\]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Design}
\label{section:Design}
In order to achieve the goal of a neural theorem prover without the disadvantages associated with neural approaches, a new design of theorem prover is required.
Popular calculi used in existing ATPs tend to be unfriendly to neural guidance.
For such a system, we require the following from the calculus:
\begin{enumerate}
        \item \emph{Proof state must be small.} Attempting to evaluate large proof states structurally requires a lot of computation and resources. Saturation-based provers can have very large proof states, for example.
        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms operate more efficiently in batches. Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space.%, without e.g. unifiers. 
        Otherwise, the learning system is trying to learn while blind to the context of the search.
	\item \emph{Subgoals must be intelligible.} Adding ``noise'' such as clausification obscures the original intuition behind a goal, at least for human observers. While this is not necessarily the case for machine-learning algorithms, it seems likely that removing structure and adding artefacts will reduce model performance.
\end{enumerate}
We therefore implement a refutation prover based on a first-order tableaux calculus without unification, on non-clausal formulae.
Each goal in this case is the set of formulae present on the tableau branch.
In this context, proof state is small (only the current branch), evaluation of states is possible in parallel, each branch is independent and contains all information required, and all available structure from the original problem is kept.

@@

Search is biased by heuristic evaluation.
The neural heuristic function (see Section \ref{section:heuristic}) evaluates each goal and assigns a score corresponding to whether the network believes that the goal is satisfiable or unsatisfiable.
In order to balance exploitation of promising directions and exploration of all parts of the search space, a principled UCT-based search algorithm is used, as in MonteCoP~\cite{MonteCoP}.
At each sub-goal $g$, the prover chooses the inference \(i\) with subgoals \(s\) according to
\[
	\max_{i \in g}\left[
		\underbrace{\min_{s \in i}\left(\textsf{score}(s)\right)}_\text{exploitation} +~
		c \times \underbrace{\sqrt{\frac{\ln\textsf{visits}\left(g\right)}{\textsf{visits}\left(i\right)}}}_\text{exploration}
	\right]
\]
where ${\sf score}$ gives the heuristic score, ${\sf visits}$ gives the total number of visits to that node so far, and \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)).
The sub-goal with the minimal score is then selected: this prioritises subgoals considered possibly satisfiable by the heuristic, as satisfiable subgoals allows large parts of the search space to be pruned.

@@

\paragraph{Refutation Tableaux}
In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and the show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
A set of inference rules of the form
\begin{mathpar}
	\inferrule{\Gamma_1 \\ \Gamma_2 \\ \ldots \\ \Gamma_n}{\Delta}
\end{mathpar}
where \(\Gamma_i, \Delta\) are sets of formulae and \(\lnot\left(\Gamma_1 \wedge \Gamma_2 \wedge \ldots \Gamma_n\right) \Rightarrow \lnot \Delta\) is an unconditional tautology, form a refutation calculus.
Proofs in this calculus can be expressed by closed trees of inference rules.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Learned Heuristic}
\label{section:heuristic}
A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatsfiability, based on a set of tagged formulae seen in previous proof search.
Although the data is collected by running the system itself and might be considered \emph{reinforcement} learning, for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.

@@

\subsection{Translation to Graphs}
Wang et al.~\cite{formula-graph} give a translation from higher-order formulae to directed graphs, and a similar scheme is used here.
Constants, function symbols, predicate symbols, and bound variables are given their own node.
Application of functions and predicates to arguments are represented as an ``application node'' with two children: the symbol node and an ``argument list'' node representing the list of arguments.
Propositional connectives and equality have the obvious representation, while quantifiers have two children: the variable they bind and their sub-formula.

To produce an input graph from a formula \(F\), the formula is first parsed into an abstract syntax tree.
Common sub-trees up to \(\alpha\)-equivalence are merged, then the resulting directed acyclic graph has any named-symbol nodes replaced with an opaque, nameless label such as "predicate" or "variable" --- since distinct symbols remain as distinct nodes under this scheme, no information is lost other than the natural-language semantics of the symbol name.
In practice, undirected graphs improved model performance so the graph is made undirected before encoding node labels as one-hot inputs to produce the final input graphs.
An example formula's translation is shown in Fig. \ref{figure:translation}.

>>>>

\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neurally-Guided, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
	Giles Reger
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}, \email{giles.reger@manchester.ac.uk}
}
\maketitle

@@

Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode the syntactic structure of proof state and therefore lose information.
By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.

@@

Processing structured data such as logical formulae is a relatively new domain for neural networks.
Some work attempts to use unstructured representations of such formulae, such as text, or build entirely-new models for a specific logic~\cite{logical-entailment}, whereas others attempt to re-use neural techniques for generic structures such as trees~\cite{logical-tree}.
A promising direction in this area is recent research on neural methods working with graphs~\cite{graph-cnn,gcn,gcn-relational}, which have already been applied to premise selection~\cite{formula-graph}.
Graph neural networks tend to include network layers inspired by convolution operators in image-processing networks, combining information from neighbouring nodes (pixels)~\cite{gcn}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Design}
\label{section:Design}
In order to achieve the goal of a neural theorem prover without the disadvantages associated with neural approaches, a new design of theorem prover is required.
Popular calculi used in existing ATPs tend to be unfriendly to neural guidance.
For such a system, we desire the following from the calculus:
\begin{enumerate}
        \item \emph{Proof state must be reasonably-sized.} Attempting to evaluate large proof states structurally requires a lot of computation and resources. Saturation-based provers can have very large proof states, for example.
        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms operate more efficiently in batches. Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space, without e.g. unifiers. 
        Otherwise, the learning system is trying to learn while blind to the context of the search.
	\item \emph{Subgoals must be intelligible.} Adding ``noise'' such as clausification obscures the original intuition behind a goal, at least for human observers. While this is not necessarily the case for machine-learning algorithms, it seems likely that removing structure and adding artefacts will reduce model performance.
\end{enumerate}
We therefore implement a refutation prover based on a first-order tableaux calculus without unification, on non-clausal formulae.
Each goal in this case is the set of formulae present on the tableau branch.
In this context, proof state is small (only the current branch), evaluation of states is possible in parallel, each branch is independent and contains all information required, and all available structure from the original problem is kept.

@@

Search is biased by heuristic evaluation.
The neural heuristic function (see Section \ref{section:heuristic}) evaluates each goal and assigns a score corresponding to whether the network believes that the goal is satisfiable or unsatisfiable.
In order to balance exploitation of promising directions and exploration of all parts of the search space, a principled UCT-based search algorithm is used, as in MonteCoP~\cite{MonteCoP}.
At each sub-goal $g$, the prover chooses the inference \(i\) with subgoals \(s\) according to
\[
	\max_{i \in g}\left[
		\underbrace{\min_{s \in i}\left(\textsf{score}(s)\right)}_\text{exploitation} +~
		c \times \underbrace{\sqrt{\frac{\ln\textsf{visits}\left(g\right)}{\textsf{visits}\left(i\right)}}}_\text{exploration}
	\right]
\]
where ${\sf score}$ gives the heuristic score, ${\sf visits}$ gives the total number of visits to that node so far, and \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)).
The sub-goal with the minimal score is then selected: this prioritises subgoals considered possibly satisfiable by the heuristic, as satisfiable subgoals allow large parts of the search space to be pruned.

@@

\paragraph{Refutation Tableaux}
In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and then show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
A set of inference rules of the form
\begin{mathpar}
	\inferrule{\Gamma_1 \\ \Gamma_2 \\ \ldots \\ \Gamma_n}{\Delta}
\end{mathpar}
where \(\Gamma_i, \Delta\) are sets of formulae and \(\lnot\left(\Gamma_1 \wedge \Gamma_2 \wedge \ldots \Gamma_n\right) \Rightarrow \lnot \Delta\) is an unconditional tautology, form a refutation calculus.
Proofs in this calculus can be expressed by closed trees of inference rules.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Learned Heuristic}
\label{section:heuristic}
A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatisfiability, based on a set of tagged formulae seen in previous proof search.
Although the data is collected by running the system itself and might be considered \emph{reinforcement} learning, for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.

@@

\subsection{Translation to Graphs}
Wang et al.~\cite{formula-graph} give a translation from higher-order formulae to directed graphs, and a similar scheme is used here.
Constants, function symbols, predicate symbols, and bound variables are given their own node.
Applications of functions and predicates to arguments are represented as an ``application node'' with two children: the symbol node and an ``argument list'' node representing the list of arguments.
Propositional connectives and equality have the obvious representation, while quantifiers have two children: the variable they bind and their sub-formula.

To produce an input graph from a formula \(F\), the formula is first parsed into an abstract syntax tree.
Common sub-trees up to \(\alpha\)-equivalence~\cite{barendregt} are merged, then the resulting directed acyclic graph has any named-symbol nodes replaced with an opaque, nameless label such as ``predicate'' or ``variable'' --- since distinct symbols remain as distinct nodes under this scheme, no information is lost other than the natural-language semantics of the symbol name.
In practice, undirected graphs improved model performance so the graph is made undirected before encoding node labels as one-hot inputs to produce the final input graphs.
An example formula's translation is shown in Fig. \ref{figure:translation}.
